Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-09-30 analysis/paper/paper.Rmd
Agrega comentarios de Leandro

---
twocol: FALSE  #  Use FALSE for submission. 
title: Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: |
  Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
  CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
  CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

@@

# Methods 
\label{sec:methods}

@@

To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.

We describe the relationship between the SAM indices and temperature and precipitation. We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude For precipitation, we use monthly, 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].

@@

Traditionally the Souther Annular Mode (SAM) is defined as de leading Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012]. Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level. We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude). We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].

To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa. The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression ofeach monthly geopotential field on the respective patterns (weighting by the cosine of latitude). The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level. As a result, the magnitude between indices is comparable. However, only "Full" index will have unit standard deviation per definition. From the regression, we also use the explained variance of each pattern as a indication of the degere of symmetry or asymmetry of each monthly field.

```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
  setnames("hgt", "full") %>% 
  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]

@@

Our method assumes linearity in the asymmetric component of the SAM. That is, we assume that zonal symmetries associated with positive SAM are oposite and equal to the ones associated with negatie SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (smaller than negative 1 standard deviation) show relatively high pattern correlations all seasons and are visually very linear (Figure A9). Therefore, we belive that our method is at the very least a reasonable approximation of the phenomenon.

@@

@@

@@

We perform linear regression to quantify the association between the SAM indices and other variables. Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions. To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the ecuation

@@

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear coefficients, $X_0$ and $\epsilon$ are the constant and error terms. From this equarion, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; in other words, it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$.

At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions. In such case, using naive p-values to test for significance leads to misleading results [@walker1914; @katz1991]. While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue. Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.

When performing a separate regression for each trimester (DJF, MAM, JJA, SON) we first average the relevant variables to obtain a single value for each year and each trimester.

# Results
\label{sec:results}

## Temporal evolution
\label{sec:temporal}

@@

Month-to-month variability is evident for both indices, with noisy variations in the low frequency. At first glance the series can be distinguished by their distributions. Compared to the stratospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound. The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.

@@

@@

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987). We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. As a consequence, statospheric Symmetric index tend to precede corresponding Asymmetric index.

@@

```{r trends, fig.cap = "Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}

indices %>% 
  copy() %>% 
  .[, season := season(time)] %>% 
  rbind(., copy(.)[, season := "Annual"]) %>%
    .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL]  %>%
  .[, estimate := estimate*3600*24*365*10] %>%
  .[, std.error := std.error*3600*24*365*10] %>%
  # .[, estimate := estimate/(3600*24*365*10)] %>%
  # .[, std.error := std.error/(3600*24*365*10)] %>%
  .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
  setnames("type", "term") %>%
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  # stat_subset(aes(subset = pval <= 0.05), shape = 4, size = 2) +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

The trends for each of the indices (Full, Symmetric, Assymetric) were evaluate for the whole period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters. The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throught the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in in Summer, where the 100 hPa maximum is much more defined. Positive trends have been documented by previous studies [e.g. @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.

By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (colum b vs. column c), but in some cases the trends become more clear. In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`). In Autumn, using the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.

@@

Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index. There is no evidencie of a significant trend in the stratosphere. In the troposphere, there is a positive trend for the Asymmetric SAM and no significant trend for the Symmetric SAM. This suggest that the SAM has become more asymmetric in the period from 1979 to 2018. The change is slight, though; of the order of 1% icreased explained variance per decade.

@@

## Spatial patterns
\label{sec:spatial}

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef. Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index. ", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>%
  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = env)) +
  geom_qmap(~.x[lat <= -20]) +
  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
                               lat = -49, term = "asym", lev = 700), 
             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL,
                   minor_breaks = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr") +
  axis_labs_smol  +
  no_grid +
  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
```

To show if, and to what extent, the Asymmetric and Symmetric SAM inidices indeed capture the asymmetric and symmetric compoment of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index. Figure \ref{fig:2d-regr} shows these regressions. Regression coefficients in column a are computed using the Full SAM. Regression coefficients in columns b and c are computed using multiple regression using the Asymmetric and Symmetric indices at the same time. Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.

In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the South Pole. The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).

In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3. The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures. The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2). The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure. Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.

@@

@@

To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}). The geopotential anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM. The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes. Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.

The tropospheric Asymmetric SAM (panel b) has significant signals that extend upwards to the uppermost levels of the reanalysis. In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa. The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere. In the Eastern hemisphere the wave-3 signal is weaker and confined to the troosphere while negative anomalies dominate in the stratosphere. So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.

>>>>

---
twocol: FALSE  #  Use FALSE for submission. 
title: Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro B. Díaz
affiliation: |
  Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
  CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
  CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

@@

# Methods

@@

To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.

To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude, and monthly rainfall 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].

@@

Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012]. Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level. We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude). We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].

To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa. The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression of each monthly geopotential field on the respective patterns (weighting by the cosine of latitude). The three indices are normalized by dividing them by the standard deviation of the "Full" index at each level. As a result, the magnitudes between indices are comparable. However, only "Full" index will have unit standard deviation per definition. From the regression, we also use the explained variance of each pattern as an indicator of the degere of zonally symmetry or asymmetry of each monthly field.

```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
  setnames("hgt", "full") %>% 
  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]

@@

Our method assumes linearity in the asymmetric component of the SAM. That is, we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negatie SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear (Figure A9). Therefore, we belive that our method is a reasonable approximation of the phenomenon.

#LEAN:  Obersvación general: Agrega en todos los captions el período que estás usando para graficar

@@

#LEAN: Deberías en el texto decir qué son las siglas de las estaciones (December-February (DJF), ...)

@@

#LEAN: ¿Se pueden referenciar bien las figuras suplementarias?

@@

We perform linear regressions to quantify the association between the SAM indices and other variables. Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions. To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the ecuation

@@

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms. From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e. it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$. When performing a separate regression for each trimester (DJF, MAM, JJA, SON), the seasonal average of the relevant variables is obtained for each year and trimester before computing the regression.

At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions. In such case, using p-values to test for significance leads to misleading results [@walker1914;@katz1991]. While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue. Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.

# Results

## Temporal evolution

@@

#LEAN: En la figura 2 quedo un poco raro ahora el "density estimate". Creo que por lo menos ayudaría que prolongues la línea del 0 para que sea más clara la lectura. Algo que también deberías agregar en la metodología es cómo se calcula.

Month-to-month variability is evident for both indices, with noisy variations in the low frequency. At first glance the series can be distinguished by their distributions. Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound. The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.

@@

#LEAN: Si va a ser la primera vez que nombra la QBO., habría que poner qué significa la sigla y la referencia correspondiente

#LEAN: Un comentario general. Creo que es mejor que siempre que hagas referencia a los indices los nombres como "SAM ..." como los nombraste en la metoodología. Si te parece muy plomo, quizás se puede poner con algun acrónimo "SAM Sym/Asym".

@@

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987). We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. As a consequence, statospheric Symmetric index tend to precede corresponding Asymmetric index.

#LEAN: Me resulto confuso en la figura 3 el caption con respecto a lo que decis. ¿Puede ser que esté al revés lo del paréntesis o yo lo estoy entendiendo mal? También me queda la duda con respecto a la última oración.

@@

```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
annual_trend <- indices %>%
  copy() %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  .[, season := "Annual"]

indices %>% 
  copy() %>% 
  # .[time > as.Date("1990-01-01")] %>%
  .[, season := season(time)] %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  rbind(annual_trend) %>% 
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
  setnames("type", "term") %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  # stat_subset(aes(subset = pval <= 0.05), shape = 4, size = 2) +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

The linear trends for each of the indices (Full, Symmetric, Assymetric) were evaluated for the complete period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters. The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throught the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined. Positive trends have been documented by previous studies [e.g. @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.

#LEAN: Esto tabień deberías incluirlo en los métodos. Cómo calculaste la tendencia (¿lineal ordinaria?) y cómo estimas el error.
#LEAN: ¿Qué pasa en invierno y primavera? Deberías indicarlo también.
#LEAN: Deberías ser más claro con la referencia en lo qué refiere a qué sucede con cada una de las estaciones y si coincide con tus resultados.

By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (colum b vs. column c), but in some cases the trends become more clear. In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`). In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index. 
#LEAN: ¿Qué pasa en invierno y primavera? Deberías indicarlo también.

@@

Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index. There is no evidencie of a significant trend in the stratosphere. In the troposphere, there is a positive trend for the Asymmetric SAM and not significant trend for the Symmetric SAM. This suggest that the SAM has become more asymmetric in the period from 1979 to 2018. However, the change is slight, around 1% increased explained variance per decade.

@@

## Spatial patterns

@@

```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices. Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index. ", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>%
  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = env)) +
  geom_qmap(~.x[lat <= -20]) +
  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
                               lat = -49, term = "asym", lev = 700), 
             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL,
                   minor_breaks = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr") +
  axis_labs_smol  +
  no_grid +
  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
```

To show if, and to what extent, the Asymmetric and Symmetric SAM inidices indeed capture the asymmetric and symmetric compoment of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels. Figure \ref{fig:2d-regr} shows these regressions. Regression coefficients in column a are computed using the Full SAM. Regression coefficients in columns b and c are computed using multiple regression using the Asymmetric and Symmetric indices at the same time. Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.

In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the South Pole. The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).

In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3. The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures. The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2). The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure. Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.

#LEAN: Deberías incluir alguna/s referencia para la primera oración

@@

#LEAN: En la metodología deberías aclarar cómo caluclas también los números de onda zonal

#LEAN: Este parráfo deberías ampliarlo un poco, primero describiendo lo que se ve en el full, las diferencias entre los niveles y luego, cómo se descompone eso entre ambas componentes.

@@

To analyse the vertical structure of the geopotential anomalies asociated with the assymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}). The geopotential anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM. The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes. Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.

#LEAN: ¿Sería geopotential height en realidad, no?

The tropospheric Asymmetric SAM (panel b) has significant signals that extend upwards to the uppermost levels of the reanalysis. In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa. The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere. In the Eastern hemisphere the wave-3 signal is weaker and confined to the troosphere while negative anomalies dominate in the stratosphere. So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM. The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are robust to the choice of index level. For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b. The patterns mainly change in amplitude (not shown).
